An Alternative Geometric Interpretation of the Riemann-Stieltjes Integral

### 2017-11-14 paper.tex
latest cav version

\begin{abstract}
It is desirable to have geometric interpretations for difficult mathematical ideas. In this paper we present an intuitive geometric interpretation of the Riemann-Stieltjes integral. It can be 
summarized as follows: the Riemann-Stieltjes integral can be interpreted as an infinite sum of infinitesimally small non-rectangular 
integration strips. This interpretation is surprisingly similar to the geometric interpretation of the ordinary Riemann integral. 
It is therefore likely that students, who are encountering the Riemann-Stieltjes integral for the first time, would find this new alternative geometric interpretation useful for mastering it.
\end{abstract}

@@

\noindent
We first present the formal definition of the Riemann-Stieltjes and Cavalieri integrals \cite{bartle1976,ackermann2012}.
We then present the geometric interpretation proposed by G.~Bullock \cite{bullock1988}, which is followed by our alternative interpretation. We end 
the paper with some examples which aim to help the reader grasp the new alternative interpretation.

@@

\noindent
Instead of trying to determine the area of $R$ using rectangular integration strips, it seems more natural 
to use integration strips whose sides are translations of $a(y)$. Cavalieri's principle tells us that the area of an 
arbitrary shaped integration strip is the product of its height and the length of its base. The area of $R$ 
can be approximated by summing together multiple non-rectangular integration strips that inscribe $R$. This concept is illustrated in Fig.~\eqref{fig:caval2}. 
More formally, consider a partition $(x_i^1)_{i=0}^{n}$, on the $x$-axis, such that $a = x_0^1 < x_1^1 < \cdots < x_n^1 = b$, and $\Delta x_i^1 = x_{i+1}^1 - x_i^1$.
%As Fig.~\ref{fig:caval2} illustrates, the partition points $x_i^1$ are chosen to coincide with the $x$-coordinates at which the sides of the 
%integration strips intersect the $x$-axis. Fig.~\ref{fig:caval2} also shows us that another partition $(x_i^2)_{i=0}^{n}$, on the $x$-axis, such that $a' = x_0^2 < x_1^2 < \cdots < x_n^2 = b'$
%is automatically created when we make use of non-rectangular integration strips. The partition points $x_i^2$ 
%coincide with the $x$-coordinates at which the sides of the 
%integration strips intersect $f(x)$. 
Let us now construct the following sum (i.e. a lower Cavalieri sum) with which we can approximate the area 
of $R$:
\begin{equation}
\label{eq:cav_sum}
\sum_{i=0}^{n-1} f(x_i^2)\Delta x_i^1.
\end{equation}
The partition points $(x_i^2)_{i=0}^{n}$ are depicted in Fig.~\ref{fig:caval2}.
The above sum can be interpreted as summing together the areas associated with $n$ non-rectangular integration strips.
The left panel of Fig.~\ref{fig:caval2} depicts the scenario in which we made use of three integration strips.  
The right panel of Fig.~\ref{fig:caval2} illustrates that the larger $n$ becomes the better Eq.~\eqref{eq:cav_sum} approximates the area of $R$. In the limit Eq.~\eqref{eq:cav_sum} approaches the Cavalieri integral \cite{ackermann2012}:
\begin{equation}
\label{eq:caval1}
\int_{a(y)}^{b(y)}f(x)\, dx = \lim_{n\to \infty}\sum_{i=0}^{n-1} f(x_i^2)\Delta x_i^1.
\end{equation}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=1.0\textwidth]{3D_plot-crop.pdf}
\caption{An example of using G.~Bullocks' geometric framework to interpret the Riemann-Stieltjes integral. In this example 
$f(x)$ is represented by the blue sheet, $g(x)$ is represented by the red sheet. Moreover, $f(x)=x^2$ and $g(x)=x$. The green sheet, simply known as the ``fence'', is formed 
by chopping of any part of the $g$-sheet which protrudes above the $f$-sheet and below the $gx$-plane. Projecting the ``fence''
onto the $fg$-plane produces the visible gray region in the $fg$-plane. The area of this gray region in the $fg$-plane is equal to the Riemann-Stieltjes integral:
$\int_{0}^{3} f(x) dg(x)$.}
\label{fig:bullock}
\end{figure}

\section{Alternative Geometric Interpretation}
\begin{figure}[htb]
\centering
\includegraphics[width=1\textwidth]{fig13_change}
\caption{A non-rectangular sum-based geometric interpretation of the Riemann-Stieltjes integral. Moreover, the figure depicts the region $R$ (also see Fig.~). The integrator $g$ maps points 
in $[a',b']$ (depicted in blue) to points in $[a,b]$ (depicted in green). The function $h$ performs the reverse operation. Moreover, $g$ maps the $x$-coordinates at which
translations of $a(y)$ (depicted in red) intersect with $f(x)$ to the $x$-intercepts of the translated functions. This observation enables us to interpret the Riemann-Stieltjes integral as an infinite sum of infinitesimally small non-rectangular integration 
strips.}
\label{fig:2d_geo}
\end{figure}
%\begin{equation}
%\label{eq:cal_def_2D}
%\int_{a(y)}^{b(y)}fdx =\lim_{n\to \infty}\sum_{i=0}^{n-1} f(x_i^2)\Delta x_i^1,
%\end{equation}
%and that
%\begin{equation}
%\label{eq:rs_def_g}
%\int_{a'}^{b'} f \, dg(x) = \lim_{n\to \infty} \sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)].
%\end{equation}
%Fig.~\ref{fig:2d_geo} depicts a region whose area can be determined by considering non-rectangular integration strips.
%. Note that the shape of the integration strips are determined by $a(y)$. 
%In Fig.~\ref{fig:2d_geo}, the function $a(y)$ is depicted in red. If we translate $a(y)$ along the $x$-axis we actually create two partitions on the $x$-axis, 
%i.e. $(x_i^1)_{i=0}^{n}$ and $(x_i^2)_{i=0}^{n}$ with $x_i^1\in[a,b]$ and $x_i^2\in[a',b']$.  To map partition points in $[a',b']$ to partition points 
%in $[a,b]$ we use the function $g$. The reverse is achieved through $h$. The functions $g$ and $h$ depend soly on $a(y)$ and are also 
%computed with it.
Consider the definition of the Riemann-Stieltjes integral and 
Eq.~\eqref{eq:g_int}. For the sake of continiuty we repeat Eq.~\eqref{eq:g_int} below:
\begin{equation}
\label{eq:g_int2}
\int_{a'}^{b'} f(x) dg(x) =  \lim_{n \rightarrow \infty}\sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)], 
\end{equation}
with $(x_i^2)_{i=0}^n$ being arbitrary partition points on the $x$-axis.
Eq.~\eqref{eq:g_int2} tells us that the function of the integrator $g$ is to map points in the interval $[a',b']$ to other 
locations on the $x$-axis. When, $g$ is monotone increasing it maps points in the interval $[a',b']$ to $[a,b]$,
with $a = g(a')$ and $b = g(b')$.\\

\noindent
To improve our understanding of $g$ lets investigate a specific example of the Riemann-Stieltjes integral in more depth: 
\begin{equation}
R = \int_{\frac{1}{2}}^1 x~d2x. 
\end{equation}
The reader is encouraged to, at this point, also inspect Fig.~\ref{fig:2d_geo} which graphically depicts $R$. 
In this example $g$ maps $[\frac{1}{2},2]$ to $[1,4]$. The mapping between these two intervals is also 
depicted in Fig.~\ref{fig:2d_geo}. Moreover, the interval $[\frac{1}{2},2]$ is depicted in blue,
while the interval $[1,4]$ is depicted in green.\\

\noindent
Furthermore, consider $a(y) = 1-y$ (depicted in red in Fig.~\ref{fig:2d_geo}). Interestingly, the integrator $g$ maps the $x$-coordinates where translations of $a(y)$ and $f(x)$ intersect to the $x$-intercepts of the translations of $a(y)$.
The implications of this observation is the following: 
\begin{itemize}
 \item $g(x_{i+1}^2)-g(x_{i}^2)$ represents the length of the base of a non-rectangular integration strip. Note, the base lengths are all equal.
 \item $f(x_i^2)$ is the height of a non-rectangular integration strip.
 \item $f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)]$ is the area of a non-rectangular integration strip (due to Cavalieri's principle).
 \item The Riemann-Stieltjes sum, i.e. $\sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)]$, is equal to the area obtained by summing together the area of multiple 
 integration strips.
 \item In the limit the Riemann-Stieltjes sum approaches the area of $R$. 
\end{itemize}
The Riemann-Stieltjes integral, can therefore, be interpreted as the area which is obtained, by summing together the areas of an infinite number of infinitesimally small non-rectangular integration strips,
i.e. a Riemann-Stieltjes integral can be converted into an equivalent Cavalieri integral.\\

@@

 \begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\textwidth]{EX1-crop.pdf}
 \caption{The non-rectangular sum-based geometric of interpretation of $\int_{1}^3 (2x+8)^3 d[x^2-20]$ (i.e this integral is equal to the 
 Cavalieri integral $\int_{a(y)}^{b(y)} (2x+8)^3 dx$).}
 \label{fig:ex1_sec}
 \end{figure}

@@

\begin{biog}
\item[Woodrow Wilson] received his Ph.D. in history and political science from Johns Hopkins University. He held visiting positions at Cornell and Wesleyan before joining the faculty at Princeton, where he was eventually appointed president of the university.  Among his proudest accomplishments was the abolition of eating clubs at Princeton on the grounds that they were elitist.
\begin{affil}
Office of the President, Princeton University, Princeton NJ 08544\\
twoodwilson@princeton.edu
\end{affil}

\item[Herbert Hoover] entered Stanford University in 1891, after failing all of the entrance exams except mathematics.  He received his B.S. degree in geology in 1895, spent time as a mining engineer, then was appointed by his co-author to the U.S. Food Administration and the Supreme Economic Council, where he orchestrated the greatest famine relief efforts of all time.
\begin{affil}
Hoover Institution, Stanford University, Stanford CA 94305\\
herbhoover@stanford.edu
\end{affil}
\end{biog}
\vfill\eject

>>>>

\begin{abstract}
It is desirable to have geometric interpretations for difficult mathematical ideas. In this paper we present an intuitive geometric interpretation of the Riemann-Stieltjes integral. It can be 
summarized as follows: the Riemann-Stieltjes integral can be interpreted as an infinite sum of infinitesimally small non-rectangular 
integration strips. This interpretation is surprisingly similar to the geometric interpretation of the ordinary Riemann integral. 
It is therefore likely that students, who are encountering the Riemann-Stieltjes integral for the first time, would find this new alternative geometric framework a useful visualization tool.
\end{abstract}

@@

\noindent
We first present the formal definitions of the Riemann-Stieltjes and Cavalieri integrals \cite{bartle1976,ackermann2012}.
We then present the geometric interpretation proposed by G.~Bullock \cite{bullock1988}, which is followed up by the new alternative interpretation. We end 
the paper with some examples which aim to help the reader better understand and visualize the new alternative interpretation.

@@

\noindent
Instead of trying to determine the area of $R$ using rectangular integration strips, it seems more natural 
to use integration strips whose sides are translations of $a(y)$. Cavalieri's principle tells us that the area of an 
arbitrary shaped integration strip is the product of its height and the length of its base. The area of $R$ 
can be approximated by summing together multiple non-rectangular integration strips that inscribe $R$. This concept is illustrated in Fig.~\eqref{fig:caval2}. 
More formally, consider a partition $(x_i^1)_{i=0}^{n}$, on the $x$-axis, such that $a = x_0^1 < x_1^1 < \cdots < x_n^1 = b$, and $\Delta x_i^1 = x_{i+1}^1 - x_i^1$.
%As Fig.~\ref{fig:caval2} illustrates, the partition points $x_i^1$ are chosen to coincide with the $x$-coordinates at which the sides of the 
%integration strips intersect the $x$-axis. Fig.~\ref{fig:caval2} also shows us that another partition $(x_i^2)_{i=0}^{n}$, on the $x$-axis, such that $a' = x_0^2 < x_1^2 < \cdots < x_n^2 = b'$
%is automatically created when we make use of non-rectangular integration strips. The partition points $x_i^2$ 
%coincide with the $x$-coordinates at which the sides of the 
%integration strips intersect $f(x)$. 
Let us now construct the following sum (i.e. a lower Cavalieri sum) with which we can approximate the area 
of $R$:
\begin{equation}
\label{eq:cav_sum}
\sum_{i=0}^{n-1} f(x_i^2)\Delta x_i^1.
\end{equation}
The partition points $(x_i^2)_{i=0}^{n}$ are depicted in Fig.~\ref{fig:caval2}.
The above sum can be interpreted as summing together the areas associated with $n$ non-rectangular integration strips.
The left panel of Fig.~\ref{fig:caval2} depicts the scenario in which we made use of three integration strips.  
The right panel of Fig.~\ref{fig:caval2} illustrates that the larger $n$ becomes the better Eq.~\eqref{eq:cav_sum} approximates the area of $R$. In the limit Eq.~\eqref{eq:cav_sum} approaches the Cavalieri integral:
\begin{equation}
\label{eq:caval1}
\int_{a(y)}^{b(y)}f(x)\, dx = \lim_{n\to \infty}\sum_{i=0}^{n-1} f(x_i^2)\Delta x_i^1.
\end{equation}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=1.0\textwidth]{3D_final-crop.pdf}
\caption{An example of using G.~Bullocks' geometric framework to interpret the Riemann-Stieltjes integral. In this example 
$f(x)$ is represented by the blue sheet, $g(x)$ is represented by the red sheet. Moreover, $f(x)=x^2$ and $g(x)=x$. The green sheet, simply known as the ``fence'', is formed 
by chopping of any part of the $g$-sheet which protrudes above the $f$-sheet and below the $gx$-plane. Projecting the ``fence''
onto the $fg$-plane produces the visible gray region in the $fg$-plane. The area of this gray region in the $fg$-plane is equal to the Riemann-Stieltjes integral:
$\int_{0}^{3} f(x) dg(x)$.}
\label{fig:bullock}
\end{figure}

\section{Alternative Geometric Interpretation}
\begin{figure}[htb]
\centering
\includegraphics[width=1\textwidth]{fig13_change}
\caption{A non-rectangular sum-based geometric interpretation of the Riemann-Stieltjes integral. Moreover, the figure depicts the region $R$ (also see Fig.~\ref{fig:ex1}). The integrator $g$ maps points 
in $[a',b']$ (depicted in blue) to points in $[a,b]$ (depicted in green). The function $h$ performs the reverse operation. Moreover, $g$ maps the $x$-coordinates at which
translations of $a(y)$ (depicted in red) intersect with $f(x)$ to the $x$-intercepts of the translated functions. This observation enables us to interpret the Riemann-Stieltjes integral as an infinite sum of infinitesimally small non-rectangular integration 
strips.}
\label{fig:2d_geo}
\end{figure}
%\begin{equation}
%\label{eq:cal_def_2D}
%\int_{a(y)}^{b(y)}fdx =\lim_{n\to \infty}\sum_{i=0}^{n-1} f(x_i^2)\Delta x_i^1,
%\end{equation}
%and that
%\begin{equation}
%\label{eq:rs_def_g}
%\int_{a'}^{b'} f \, dg(x) = \lim_{n\to \infty} \sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)].
%\end{equation}
%Fig.~\ref{fig:2d_geo} depicts a region whose area can be determined by considering non-rectangular integration strips.
%. Note that the shape of the integration strips are determined by $a(y)$. 
%In Fig.~\ref{fig:2d_geo}, the function $a(y)$ is depicted in red. If we translate $a(y)$ along the $x$-axis we actually create two partitions on the $x$-axis, 
%i.e. $(x_i^1)_{i=0}^{n}$ and $(x_i^2)_{i=0}^{n}$ with $x_i^1\in[a,b]$ and $x_i^2\in[a',b']$.  To map partition points in $[a',b']$ to partition points 
%in $[a,b]$ we use the function $g$. The reverse is achieved through $h$. The functions $g$ and $h$ depend soly on $a(y)$ and are also 
%computed with it.
Consider the definition of the Riemann-Stieltjes integral and 
Eq.~\eqref{eq:g_int}. For the sake of continiuty we repeat Eq.~\eqref{eq:g_int} below:
\begin{equation}
\label{eq:g_int2}
\int_{a'}^{b'} f(x) dg(x) =  \lim_{n \rightarrow \infty}\sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)], 
\end{equation}
with $(x_i^2)_{i=0}^n$ being arbitrary partition points on the $x$-axis.
Eq.~\eqref{eq:g_int2} tells us that the function of the integrator $g$ is to map points in the interval $[a',b']$ to other 
locations on the $x$-axis. When, $g$ is monotone increasing it maps points in the interval $[a',b']$ to $[a,b]$,
with $a = g(a')$ and $b = g(b')$.\\

\noindent
To improve our understanding of $g$ lets investigate a specific example of the Riemann-Stieltjes integral in more depth: 
\begin{equation}
R = \int_{\frac{1}{2}}^1 x~d2x. 
\end{equation}
$R$ is depicted in Fig.~\ref{fig:2d_geo}. In this example $g$ maps $[\frac{1}{2},2]$ to $[1,4]$. The mapping between these two intervals is also 
depicted in Fig.~\ref{fig:2d_geo}. Moreover, the interval $[\frac{1}{2},2]$ is depicted in blue,
while the interval $[1,4]$ is depicted in green.\\

\noindent
Furthermore, consider $a(y) = 1-y$ (depicted in red in Fig.~\ref{fig:2d_geo}). Interestingly, the integrator $g$ maps the $x$-coordinates where translations of $a(y)$ and $f(x)$ intersect to the $x$-intercepts of the translations of $a(y)$.
The implications of this observation is the following: 
\begin{itemize}
 \item $g(x_{i+1}^2)-g(x_{i}^2)$ represents the length of the base of a non-rectangular integration strip. Note, the base lengths in Fig.~\ref{fig:2d_geo} are all equal.
 \item $f(x_i^2)$ is the height of a non-rectangular integration strip.
 \item $f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)]$ is the area of a non-rectangular integration strip (due to Cavalieri's principle).
 \item The Riemann-Stieltjes sum, i.e. $\sum_{i=0}^{n-1} f(x_i^2)[g(x_{i+1}^2)-g(x_{i}^2)]$, is equal to the area obtained by summing together the area of multiple 
 integration strips.
 \item In the limit the Riemann-Stieltjes sum approaches $R$. 
\end{itemize}
The Riemann-Stieltjes integral, can therefore, be interpreted as the area which is obtained, by summing together the areas of an infinite number of infinitesimally small non-rectangular integration strips,
i.e. a Riemann-Stieltjes integral can be converted into an equivalent Cavalieri integral.\\

@@

 \begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\textwidth]{EX1-crop.pdf}
 \caption{The non-rectangular sum-based geometric interpretation of $\int_{1}^3 (2x+8)^3 d[x^2-20]$ (i.e this integral is equal to the 
 Cavalieri integral $\int_{a(y)}^{b(y)} (2x+8)^3 dx$).}
 \label{fig:ex1_sec}
 \end{figure}

@@

\begin{biog}
\item[Trienko Lups Grobler] received his Ph.D. in Engineering at the University of Pretoria in 2013. He was 
a postdoctoral fellow at Rhodes University from 2013-2017. He is currently a lecturer in the Computer Science Division
at Stellenbosch University.
\begin{affil}
Dept of Mathematical Sciences, Computer Science Division, Stellenbosch University, Private Bag X1, 7602 Matieland, South Africa\\
tlgrobler@sun.ac.za
\end{affil}

%\item[Herbert Hoover] entered Stanford University in 1891, after failing all of the entrance exams except mathematics.  He received his B.S. degree in geology in 1895, spent time as a mining engineer, then was appointed by his co-author to the U.S. Food Administration and the Supreme Economic Council, where he orchestrated the greatest famine relief efforts of all time.
%\begin{affil}
%Hoover Institution, Stanford University, Stanford CA 94305\\
%herbhoover@stanford.edu
%\end{affil}
\end{biog}
\vfill\eject
